---
title: 'Momentary excitability of human PFC estimated by EEG aperiodic component predicts pain intensity and the size of pain-evoked brain response'
subtitle: 'Statistics and visualization'
author: 'Dominika Sulcova'
institute: 'MSH - Medical School Hamburg'
date: "`r Sys.Date()`"
output: pdf_document
---
```{r libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(cowplot)
library(ggpubr)
library(ggforce)
library(reshape2)
#library(eegUtils)
library(gridExtra)
library(knitr)
library(rmarkdown)
library(rlang)
library(svglite)
library(car)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 50),tidy = TRUE)
```

```{r,message=FALSE,warning=FALSE}
# set up the working directory
wd_path <- choose.dir(default = getwd(), caption = "Select a Folder")
if (nzchar(wd_path)) {
  setwd(wd_path)
  message("Working directory set to: ", wd_path)
} else {
  message("No folder selected. Working directory remains: ", getwd())
}

# functions
summarize_and_plot <- function(data, variable, ..., group_vars = NULL) {
  # convert variables to symbols
  variable <- sym(variable)
  variable_name <- quo_name(variable)
  if (!is.null(group_vars)) {
    group_vars <- syms(group_vars)
  }
  
  # determine variable
  if (grepl("amplitude", variable_name, ignore.case = TRUE)) {
    y_label <- "amplitude (µV)"
  } else if (grepl("latency", variable_name, ignore.case = TRUE)) {
    y_label <- "latency (s)"
  } else if (grepl("pain", variable_name, ignore.case = TRUE)) {
    y_label <- "intensity (% VAS)"
  } else if (grepl("slope", variable_name, ignore.case = TRUE)) {
    y_label <- "slope"
  } else {
    y_label <- "value"
  }
  
  custom_theme <- theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.9)
  )
  
  # calculate summary statistics
  if (length(group_vars) == 0) {
    summary_stats <- data %>%
      summarise(
        mean = mean(!!variable, na.rm = TRUE),
        sd = sd(!!variable, na.rm = TRUE),
        sem = sd / sqrt(n())
      )
  } else {
    summary_stats <- data %>%
      group_by(!!!group_vars) %>%
      summarise(
        mean = mean(!!variable, na.rm = TRUE),
        sd = sd(!!variable, na.rm = TRUE),
        sem = sd / sqrt(n())
      ) %>%
      ungroup()
  }
  
  # print summary statistics
  print(summary_stats)
  
  # plot 1: distribution of all values
  dist_plot <- ggplot(data, aes(x = !!variable)) +
    geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
    labs(title = paste(variable_name, ": distribution of values"), x = y_label, y = "frequency") + 
    custom_theme
  
  # plot 2: boxplot with single-subject averages
  subject_averages <- data %>%
    group_by(subject) %>%
    summarise(
      subject_mean = mean(!!variable, na.rm = TRUE),
      subject_sd = sd(!!variable, na.rm = TRUE)) %>%
    ungroup()
  
  grand_mean <- mean(subject_averages$subject_mean, na.rm = TRUE)
  grand_sd <- sd(subject_averages$subject_mean, na.rm = TRUE)
  
  dot_plot <- ggplot(subject_averages, aes(x = factor(subject), y = subject_mean)) +
    geom_point(color = "blue", size = 3) +
    geom_errorbar(aes(ymin = subject_mean - subject_sd, ymax = subject_mean + subject_sd), width = 0.3) +
    geom_hline(yintercept = grand_mean, linetype = "dashed", color = "red", width = 1.2) +
    annotate("rect", xmin = -Inf, xmax = Inf, ymin = grand_mean - grand_sd, ymax = grand_mean + grand_sd, 
             alpha = 0.2, fill = "red") +
    labs(title = paste(variable_name, ": average values per subject"), x = 'Subject', y = y_label) +
    custom_theme
  
  # additional plots if grouping variables are provided
  if (!is.null(group_vars)) {
    # distribution plot for each group
    dist_plot_group <- ggplot(data, aes(x = !!variable, fill = factor(!!!group_vars))) +
      geom_histogram(position = "identity", bins = 30, alpha = 0.4) +
      labs(title = "Distribution of values by group", x = y_label, y = "frequency") +
      theme(legend.title = element_blank()) +
      custom_theme
    
    # boxplot for each group with subject markers
    boxplot_group <- ggplot(data, aes(x = interaction(!!!group_vars), y = !!variable)) +
      geom_jitter(aes(color = factor(subject)), width = 0.2, height = 0, alpha = 0.6) +
      geom_boxplot(outlier.shape = NA, alpha = 0.2, fill = NA) +
      stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "red") +
      labs(title = paste(variable_name, ": mean group values"), x = paste(group_vars, collapse = " "), y = y_label) +
      theme(legend.position = "none") +
      custom_theme
    if (grepl("latency", variable_name, ignore.case = TRUE)) {
      boxplot_group <- boxplot_group + coord_flip()
      }
    
    return(list(summary_stats = summary_stats, dist_plot = dist_plot, dot_plot = dot_plot,
            dist_plot_group = dist_plot_group, boxplot_group = boxplot_group))
    }
  
  # display the plots
  print(dist_plot)
  print(dot_plot)
  if (!is.null(group_vars)) {
    print(dist_plot_group)
    print(boxplot_group)
  }
  
  # return summary statistics and plot handles
  return(list(summary_stats = summary_stats, dist_plot = dist_plot, dot_plot = dot_plot))
}
check_model <- function(lmm, data){
  # visualize residuals
  plot(fitted(lmm),residuals(lmm))
  qqnorm(residuals(lmm))
  qqline(residuals(lmm))
  ggplot(data.frame(cook=cooks.distance(lmm),id=data$subject),
         aes(x=cook,y=id)) +
    geom_point() +
    theme_bw()
  
  # identify outliers using the 1.5 * IQR rule
  residuals <- resid(lmm)
  residuals_df <- data.frame(residuals = residuals)
  IQR_res <- IQR(residuals_df$residuals)
  Q1 <- quantile(residuals_df$residuals, 0.25)
  Q3 <- quantile(residuals_df$residuals, 0.75)
  lower_bound <- Q1 - 1.5 * IQR_res
  upper_bound <- Q3 + 1.5 * IQR_res
  outliers <- residuals_df %>%
  filter(residuals < lower_bound | residuals > upper_bound)
  
  # return outliers
  return(outliers)
}  
```

This script provides the basis for statistical analysis and visualization of data collected in 2024 in **Laserlab at MSH Medical School Hamburg**. The aim was to obtain a comprehensive dataset of normative LEPs from young healthy population and to use this data to test several hypotheses. **LEPs** were recorded wih 64-channel Brain Vision system (original SR 5kHz, online ref FCz) following stimulation with Yap laser with fixed parameters (1.75J, 5mm, 3ms). Each subject was stimulated on **two body areas**, either on both hands, both feet, or on a hand and a foot on the same side. Stimuli were delivered in blocks of 30 in ABBA format, resulting in block 1 and 2 for each area. In addition, each subject underwent psychological evaluation using questionnaires (BDI, PCS), evaluation of sensory system using shortened version of QST, and recording of **SEPs** using electrical stimulation of either median or tibial nerve (300 stimuli delivered to the same areas and in the same order). **Resting-state EEG** was obtained in 1.5 min blocks with eyes open and closed before the LEP recording and after the first two blocks of laser stimulation. 

Here, the collected data is used to investigate, whether the momentary excitability of the prefrontal cortex (PFC) can predict the character of brain response to painful stimulation and the subjective pain intensity. PFC excitability was estimated from 1s segments of EEG recording preceding the laser stimulus obtained at the frontal electrodes. More specifically, the spectra of single-trial signals were separated into the oscillatory and aperiodic component and the **slope of the aperiodic component (`r if (knitr::is_html_output()) '&beta;' else '$\\beta$'`)** was cosidered as an approximation of the ratio of excitatory and inhibitory neurotransmission. The **amplitude and latency of LEP components N1, N2 and P2** were also obtained at single-trial level, and the **intensity of perceived pain** was encoded as a rating provided by the subject after each stimulus using a VAS (0-100). The association between the predictor variable (`r if (knitr::is_html_output()) '&beta;' else '$\\beta$'` parameter) and each outcome variable was tested using LMMs.     

## DATA IMPORT
```{r,message=FALSE,warning=FALSE}
# read .csv files
data_all <- read_csv(file.path(wd_path, paste0('AperiodicPFC_export_sources_hemisphere.csv')))
data2exclude <- read_csv2(file.path(wd_path, 'AperiodicPFC_data2exclude.csv'))
data2exclude <- data2exclude %>%
  filter(!is.na(subject))

# explore the structure of data and summarize 
str(data_all)
summary(data_all)

# clean and prepare complete dataset
data_all <- data_all %>%
  mutate(across(where(~!is.numeric(.)), as.factor)) %>% 
  filter(latency > 0) %>% 
  mutate(region = if_else(region == "Vis_LH", "Visual_LH", region)) %>% 
  mutate(region = if_else(region == "Vis_RH", "Visual_RH", region)) %>% 
  mutate(brain_area = if_else(brain_area == "Vis", "Visual", brain_area))
data_all <- data_all %>%
  group_by(subject) %>%
  mutate(
    rating_norm = (rating / max(rating, na.rm = TRUE)) * 100
  ) %>%
  ungroup()
exclude <- list()
for(i in 1:nrow(data2exclude)) {
  # extract the current exclusion criteria
  exclude$subj <- data2exclude$subject[i]
  exclude$area <- data2exclude$area[i]
  exclude$side <- data2exclude$side[i]
  exclude$comp <- data2exclude$component[i]
  
  if(exclude$comp == "all") {
    # exclude all components for this subject/area/side
    data_all <- data_all %>%
      filter(!(subject == exclude$subj & area == exclude$area & side == exclude$side))
  } else {
    # exclude specific component
    data_all <- data_all %>%
      filter(!(subject == exclude$subj & area == exclude$area & side == exclude$side & component == exclude$comp))
  }
}
rm(exclude, i)
```

## DESCRIPTIVE STATISTICS
# subjects
```{r}
data_subjects <- data_all %>%
  # group_by(subject, area, component) %>%
  group_by(subject) %>%
  summarize(
    age = first(age),  
    male = first(male),
    handedness = first(handedness)
  )

# data_subj <- data_subjects %>%
#   filter(area == 'foot') %>% 
#   filter(component == 'N2')

summary_stats <- data_subjects %>%
  summarize(
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    num_males = sum(male == 1, na.rm = TRUE),
    num_righthanded = sum(handedness == 'right', na.rm = TRUE)
  )
print(summary_stats)
rm(data_subjects, data_subj)
```

# LEPs
```{r}
# define variables
components <- c('N1', 'N2', 'P2')
areas <- c('hand', 'foot')
variables <- c('amplitude', 'latency')

# prepare outcome df
outcome_LEPs <- data.frame(component = character(),
                         area = character(),
                         variable = character(),
                         mean = numeric(),
                         sd = numeric(),
                         sem = numeric(),
                         stringsAsFactors = FALSE)

# loop through components, areas, and variables 
for (comp in components) {
  for (ar in areas) {
    for (var in variables) {
      # filter the data for the current component and area
      data_LEPs <- data_all %>% 
        filter(component == !!comp) %>% 
        filter(area == !!ar) 
      
      # summarize and plot
      output <- summarize_and_plot(data_LEPs, var)
      
      # append the results to the outcome variable
      outcome_LEPs <- outcome_LEPs %>%
        add_row(component = comp,
                area = ar,
                variable = var,
                mean = output$summary_stats$mean,
                sd = output$summary_stats$sd,
                sem = output$summary_stats$sem)
    
      # save the plots as SVG files
      save_plot(paste0(wd_path, '\\figures\\', comp, "_", ar, "_", var, "_distribution.svg"), output$dist_plot)
      save_plot(paste0(wd_path, '\\figures\\', comp, "_", ar, "_", var, "_subjects.svg"), output$dot_plot)
    }
  }
}
print(outcome_LEPs)

# loop through areas and variables, group by components
for (ar in areas) {
  for (var in variables) {
      # filter the data for the current peak
      data_LEPs <- data_all %>% 
        filter(area == !!ar) 
      
      # summarize and plot
      output <- summarize_and_plot(data_LEPs, var, group_vars = 'component')
      
      # display distribution
      print(output$dist_plot_group)
      save_plot(paste0(wd_path, '\\figures\\LEP_', ar, '_', var, "_distribution_all.svg"), output$dist_plot_group)
      
      # display boxplot
      print(output$boxplot_group)
      save_plot(paste0(wd_path, '\\figures\\LEP_', ar, '_', var, "_boxplot_all.svg"), output$boxplot_group)
  }
}
rm(components, areas, variables, comp, ar, var, data_LEPs, output)
```

# pain ratings
```{r}
# define variables
areas <- c('hand', 'foot')
variables <- c('rating', 'rating_norm')

# prepare outcome df
outcome_ratings <- data.frame(area = character(),
                              variable = character(),
                              mean = numeric(),
                              sd = numeric(),
                              sem = numeric(),
                              stringsAsFactors = FALSE)

# loop through areas
for (ar in areas) {
  for (var in variables) {
    # filter the data for the current area
    data_ratings <- data_all %>% 
      filter(area == !!ar) 
  
    #  summarize and plot
    output <- summarize_and_plot(data_ratings, var)
    save_plot(paste0(wd_path, '\\figures\\', var, '_', ar, '_distribution.svg'), output$dist_plot)
    save_plot(paste0(wd_path, '\\figures\\', var, '_', ar, '_subjects.svg'), output$dot_plot)
    
    # append the results to the outcome variable
    outcome_ratings <- outcome_ratings %>%
      add_row(area = ar,
              variable = var,
              mean = output$summary_stats$mean,
              sd = output$summary_stats$sd,
              sem = output$summary_stats$sem)
  }
}

# loop through variables, group by areas
for (var in variables) {
    # summarize and plot
    output <- summarize_and_plot(data_all, var, group_vars = 'area')
    
    # display distribution
    print(output$dist_plot_group)
    save_plot(paste0(wd_path, '\\figures\\', var, '_distribution_all.svg'), output$dist_plot_group)
    
    # display boxplot
    print(output$boxplot_group)
    save_plot(paste0(wd_path, '\\figures\\', var, "_boxplot_all.svg"), output$boxplot_group)
}
rm(areas, variables, ar, var, data_ratings, output)
```

# aperiodic measures
```{r}
# define variables
areas <- c('hand', 'foot')
regions <- c('PFC_LH', 'PFC_LH', 'Visual_LH', 'Visual_LH')
variables <- c('exponent', 'offset')

# prepare outcome df
outcome_APC <- data.frame(area = character(),
                          region = character(),
                          variable = character(),
                          mean = numeric(), 
                          sd = numeric(),
                          sem = numeric(),
                          stringsAsFactors = FALSE)

# loop through areas, regions, and variables 
for (ar in areas) {
  for (reg in regions) {
    for (var in variables) {
      # filter the data for the current area and brain region
      data_APC <- data_all %>% 
        filter(area == !!ar) %>%
        filter(region == !!reg) 
      
      # summarize and plot
      output <- summarize_and_plot(data_APC, var)
      
      # append the results to the outcome variable
      outcome_APC <- outcome_APC %>%
        add_row(area = ar,
                region = reg,
                variable = var,
                mean = output$summary_stats$mean,
                sd = output$summary_stats$sd,
                sem = output$summary_stats$sem)
    
      # save the plots as SVG files
      # save_plot(paste0(wd_path, '\\figures\\APC-source_', ar, '_', var, "_", reg, "_distribution.svg"), output$dist_plot)
      # save_plot(paste0(wd_path, '\\figures\\APC-source_', ar, '_', var, "_", reg, "_subjects.svg"), output$dot_plot)
    }
  }
}
print(outcome_APC)

# loop through areas and variables, group by regions
for (ar in areas) {
  for (var in variables) {
      # filter the data for the current peak
      data_APC <- data_all %>%
        filter(area == !!ar)
      
      # summarize and plot
      output <- summarize_and_plot(data_APC, var, group_vars = 'region')
      # output <- summarize_and_plot(data_all, var, group_vars = 'region')
      
      # display distribution
      print(output$dist_plot_group)
      # save_plot(paste0(wd_path, '\\figures\\APC-source_', ar, '_', var,  "_distribution_all.svg"), output$dist_plot_group)
      
      # display boxplot
      print(output$boxplot_group)
      # save_plot(paste0(wd_path, '\\figures\\APC-source_', ar, '_', var,  "_boxplot_all.svg"), output$boxplot_group)
  }
}

# loop through variables, group by regions
for (var in variables) {
    # summarize and plot
    output <- summarize_and_plot(data_all, var, group_vars = 'region')
    
    # display distribution
    print(output$dist_plot_group)
    # save_plot(paste0(wd_path, '\\figures\\APC-source_avg_', var,  "_distribution_all.svg"), output$dist_plot_group)
    
    # display boxplot
    print(output$boxplot_group)
    # save_plot(paste0(wd_path, '\\figures\\APC-source_avg_', var,  "_boxplot_all.svg"), output$boxplot_group)
}
rm(areas, regions, variables, ar, reg, var, data_APC, output)
```

## STATISTICAL ANALYSIS
# N1 amplitude
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N1") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'region', 'amplitude', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(amplitude ~ exponent + offset + region + area, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(amplitude ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(amplitude ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(amplitude ~ exponent + region + area + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(amplitude ~ exponent * region + area + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# apply log-transformation to the outcome variable
min_value <- min(data$amplitude, na.rm = TRUE)
data <- data %>%
  mutate(amplitude_shifted = amplitude - min_value + 1,
    amplitude_log = log(amplitude_shifted))
model_log <- lmer(amplitude_log ~ exponent * region + area + (1 | subject),
  data = data)
anova(model_log)
anova_table <- anova(model_log)
# view(anova_table)
outliers <- check_model(model_log, data)
```

# N1 latency
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N1") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'region', 'latency', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(latency ~ exponent + offset + region + area, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(latency ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(latency ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(latency ~ exponent + region + area + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(latency ~ exponent * region + area + (1 | subject), data = data)
anova(model_interaction)
anova_table <- anova(model_interaction)
# view(anova_table)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)
```

# N2 amplitude
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N2") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'region', 'amplitude', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(amplitude ~ exponent + offset + region + area, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(amplitude ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(amplitude ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(amplitude ~ exponent + region + area + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(amplitude ~ exponent * region + area + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# apply log-transformation to the outcome variable
min_value <- min(data$amplitude, na.rm = TRUE)
data <- data %>%
  mutate(amplitude_shifted = amplitude - min_value + 1,
    amplitude_log = log(amplitude_shifted))
model_log <- lmer(amplitude_log ~ exponent * region + area + (1 | subject),
  data = data)
anova(model_log)
anova_table <- anova(model_log)
# view(anova_table)
outliers <- check_model(model_log, data)
```

# N2 latency
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N2") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'region', 'latency', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(latency ~ exponent + offset + region + area, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(latency ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(latency ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(latency ~ exponent + region + area + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(latency ~ exponent * region + area + (1 | subject), data = data)
anova(model_interaction)
anova_table <- anova(model_interaction)
# view(anova_table)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)
```

# P2 amplitude
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "P2") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'region', 'amplitude', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(amplitude ~ exponent + offset + region + area, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(amplitude ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(amplitude ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(amplitude ~ exponent + region + area + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(amplitude ~ exponent * region + area + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# apply log-transformation to the outcome variable
min_value <- min(data$amplitude, na.rm = TRUE)
data <- data %>%
  mutate(amplitude_shifted = amplitude - min_value + 1,
    amplitude_log = log(amplitude_shifted))
model_log <- lmer(amplitude_log ~ exponent * region + area + (1 | subject),
  data = data)
anova(model_log)
anova_table <- anova(model_log)
# view(anova_table)
outliers <- check_model(model_log, data)
```

# P2 latency
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "P2") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'region', 'latency', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(latency ~ exponent + offset + region + area, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(latency ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(latency ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(latency ~ exponent + region + area + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(latency ~ exponent * region + area + (1 | subject), data = data)
anova(model_interaction)
anova_table <- anova(model_interaction)
# view(anova_table)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)
```

# pain rating
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N2") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'brain_area', 'region', 'rating', 'rating_norm', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(rating ~ exponent + offset + region + area, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(rating ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(rating ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(rating ~ exponent + region + area + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(rating ~ exponent * region + area + (1 | subject), data = data)
anova(model_interaction)
anova_table <- anova(model_interaction)
# view(anova_table)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# try computing a separate model for the brain areas
data_PFC <- data %>% 
  filter(brain_area == "PFC")
model_PFC <- lmer(rating ~ exponent * region + area + (1 | subject), data = data_PFC)
anova(model_PFC)
outliers <- check_model(model_PFC, data_PFC)

data_visual <- data %>% 
  filter(brain_area == "Visual")
model_visual <- lmer(rating ~ exponent * region + area + (1 | subject), data = data_visual)
anova(model_visual)
outliers <- check_model(model_visual, data_visual)

# try computing a separate model for the stimulated areas
data_hand <- data %>% 
  filter(area == "hand")
model_hand <- lmer(rating ~ exponent*region + (1 | subject), data_hand)
anova(model_hand)
anova_table <- anova(model_hand)
view(anova_table)
outliers <- check_model(model_hand, data_hand)

data_foot <- data %>% 
  filter(area == "foot")
model_foot <- lmer(rating ~ exponent*region + (1 | subject), data_foot)
anova(model_foot)
anova_table <- anova(model_foot)
view(anova_table)
outliers <- check_model(model_foot, data_foot)

```

## STATISTICAL ANALYSIS - only hand stimulation
# N1 amplitude
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N1") %>% 
  filter(area == "hand") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'region', 'amplitude', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(amplitude ~ exponent + offset + region, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(amplitude ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(amplitude ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(amplitude ~ exponent + region + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(amplitude ~ exponent * region + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# apply log-transformation to the outcome variable
min_value <- min(data$amplitude, na.rm = TRUE)
data <- data %>%
  mutate(amplitude_shifted = amplitude - min_value + 1,
    amplitude_log = log(amplitude_shifted))
model_log <- lmer(amplitude_log ~ exponent * region + (1 | subject),
  data = data)
anova(model_log)
anova_table <- anova(model_log)
view(anova_table)
outliers <- check_model(model_log, data)
```

# N1 latency
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N1") %>% 
  filter(area == "hand") %>% 
  select(c('subject', 'age', 'component', 'dominant', 'area', 'region', 'latency', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(latency ~ exponent + offset + region, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(latency ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(latency ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(latency ~ exponent + region + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(latency ~ exponent * region + (1 | subject), data = data)
anova(model_interaction)
anova_table <- anova(model_interaction)
view(anova_table)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# try computing a separate model for the regions
data_target <- data %>% 
  filter(region == "target")
model_target <- lmer(latency ~ exponent + (1 | subject), data = data_target)
anova(model_target)
outliers <- check_model(model_target, data_target)

data_ctrl <- data%>% 
  filter(region == "ctrl")
model_ctrl <- lmer(latency ~ exponent + (1 | subject), data = data_ctrl)
anova(model_ctrl)
outliers <- check_model(model_ctrl, data_ctrl)
```

# N2 amplitude
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N2") %>% 
  filter(area == "hand") %>% 
  select(c('subject', 'age', 'male', 'dominant', 'region', 'amplitude', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(amplitude ~ exponent + offset + region, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(amplitude ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(amplitude ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(amplitude ~ exponent + region + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(amplitude ~ exponent * region + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# apply log-transformation to the outcome variable
min_value <- min(data$amplitude, na.rm = TRUE)
data <- data %>%
  mutate(amplitude_shifted = amplitude - min_value + 1,
    amplitude_log = log(amplitude_shifted))
model_log <- lmer(amplitude_log ~ exponent * region + (1 | subject),
  data = data)
anova(model_log)
outliers <- check_model(model_log, data)
```

# N2 latency
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N2") %>% 
  filter(area == "hand") %>% 
  select(c('subject', 'age', 'male', 'dominant', 'area', 'region', 'latency', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(latency ~ exponent + offset + region, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(latency ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(latency ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(latency ~ exponent + region + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(latency ~ exponent * region + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# try computing a separate model for the regions
data_target <- data %>% 
  filter(region == "target")
model_target <- lmer(latency ~ exponent + (1 | subject), data = data_target)
anova(model_target)
outliers <- check_model(model_target, data_target)

data_ctrl <- data%>% 
  filter(region == "ctrl")
model_ctrl <- lmer(latency ~ exponent + (1 | subject), data = data_ctrl)
anova(model_ctrl)
outliers <- check_model(model_ctrl, data_ctrl)

```

# P2 amplitude
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "P2") %>% 
  filter(area == "hand") %>% 
  select(c('subject', 'age', 'male', 'dominant', 'region', 'amplitude', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(amplitude ~ exponent + offset + region, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(amplitude ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(amplitude ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(amplitude ~ exponent + region + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(amplitude ~ exponent * region + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# apply log-transformation to the outcome variable
min_value <- min(data$amplitude, na.rm = TRUE)
data <- data %>%
  mutate(amplitude_shifted = amplitude - min_value + 1,
    amplitude_log = log(amplitude_shifted))
model_log <- lmer(amplitude_log ~ exponent * region + (1 | subject),
  data = data)
anova(model_log)
outliers <- check_model(model_log, data)
```

# P2 latency
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "P2") %>% 
  filter(area == "hand") %>% 
  select(c('subject', 'age', 'male', 'dominant', 'area', 'region', 'latency', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(latency ~ exponent + offset + region, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(latency ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(latency ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(latency ~ exponent + region + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(latency ~ exponent * region + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# try computing a separate model for the regions
data_target <- data %>% 
  filter(region == "target")
model_target <- lmer(latency ~ exponent + (1 | subject), data = data_target)
anova(model_target)
outliers <- check_model(model_target, data_target)

data_ctrl <- data%>% 
  filter(region == "ctrl")
model_ctrl <- lmer(latency ~ exponent + (1 | subject), data = data_ctrl)
anova(model_ctrl)
outliers <- check_model(model_ctrl, data_ctrl)
```

# pain rating
```{r}
# subset the dataset
data <- data_all %>%
  filter(component == "N2") %>% 
  filter(area == "hand") %>% 
  select(c('subject', 'age', 'male', 'dominant', 'area', 'region', 'rating', 'rating_norm', 'exponent', 'offset')) 

# check for multicollinearity predictors --> Variance Inflation Factors (VIF)
lm_check <- lm(rating ~ exponent + offset + region, data = data)
vif(lm_check)
# ==> We see very high collinearity between the aperiodic exponent and offset. We will therefore not add offset as a covariate.

# empty model
model_null <- lmer(rating ~ 1 + (1 | subject), data = data)
anova(model_null)
outliers <- check_model(model_null, data)

# basic model = only with exponent 
model_basic <- lmer(rating ~ exponent + (1 | subject), data = data)
anova(model_basic)
outliers <- check_model(model_basic, data)

# full model = with main covariates
model_full <- lmer(rating ~ exponent + region + (1 | subject), data = data)
anova(model_full)
outliers <- check_model(model_full, data)

# interaction model = includes an interaction between exponent and brain region
model_interaction <- lmer(rating ~ exponent * region + (1 | subject), data = data)
anova(model_interaction)
outliers <- check_model(model_interaction, data)

# compare models
anova(model_null, model_basic, model_full, model_interaction)
AIC(model_null, model_basic, model_full, model_interaction)

# try computing a separate model for the regions
data_lateral <- data %>% 
  filter(region == "lateral")
model_lateral <- lmer(rating ~ exponent + (1 | subject), data = data_lateral)
anova(model_lateral)
outliers <- check_model(model_lateral, data_lateral)

data_medial <- data %>% 
  filter(region == "medial")
model_medial <- lmer(rating ~ exponent + (1 | subject), data = data_medial)
anova(model_medial)
outliers <- check_model(model_medial, data_medial)

data_visual <- data %>% 
  filter(region == "visual")
model_visual <- lmer(rating ~ exponent + (1 | subject), data = data_visual)
anova(model_visual)
outliers <- check_model(model_visual, data_visual)
```
